How to improve your relationship with your future self.

### 2016-06-24 workflow.Rmd
some more edits
ok ready for a discussion with you

---
title: How to improve your relationship with your future self.
author:
- name: Jake Bowers
  affiliation: <jwbowers@illinois.edu>,  University of Illinois at Urbana-Champaign, Departments of Political Science and Statistics; Fellow, White House Social and Behavioral Sciences Team. Many thanks to Mark Fredrickson, Brian Gaines, Kieran Healy, Kevin Quinn and Cara Wong for direct help on this document and to Mika LaVaque-Manty and Ben Hansen for many useful discussions on this topic. The source code for this document may be freely downloaded and modified from <https://github.com/jwbowers/workflow>.
- name: Maarten Voors
  affiliation: <maarten.voors@wur.nl>, Wageningen University, Development Economics Group
bibliography: references.bib
csl: american-political-science-review.csl
date: '`r format(Sys.Date(), "%B %d, %Y")`'
output:
    pdf_document:
      keep_tex: true
      number_sections: true
      fig_width: 7
      fig_height: 7
      fig_caption: true
      template: bowersarticletemplate.latex
      pandoc_args: [
      "-M", "secPrefix=\\S",
      "--filter", "/usr/local/bin/pandoc-crossref"
      ]
      md_extensions: +autolink_bare_uris+ascii_identifiers+tex_math_single_backslash
...

@@

Memory is tricky. When we really want to remember something we should practice it until it becomes internalized. In fact, learning requires effort. When we do not practice, we tend to be overconfident in our recall abilities [@koriat2005illusions] and false memories can be implanted.^[See the following site for a nice overview of what we know about memory including the fact that learning requires practice: <http://www.spring.org.uk/2012/10/how-memory-works-10-things-most-people-get-wrong.php>. On false memory see Wikipedia and linked studies <https://en.wikipedia.org/wiki/False_memory>.] If we cannot count on memory, then how can we do science?

How long does it take from planning a study to publication? And subsequently, how much time passes to the first reproduction of a study after publication? Is three years too short? Is ten years too long? We suspect that few of our colleagues in the social and behavioral sciences conceive of, field, write, and publish a study in faster than about three years. We also suspect that, if some other scholar decides to re-examine the analyses of a published study, it will occur after publication, and that this new scholarly activity of learning from one anothers' data and analyses, can occur at anytime, years past the initial publication of the article.^[We note that sometimes this reproduction occurs within the context of classes and learning about statistics and data analysis. Other times, this process is organized more broadly (cite IIIE program, cite Open Science Foundation on psychology, cite the Camerer 2016 Science paper?, cite the cool "Crowdsourcing Analysis" project by Silberzahn and Uhlmann Nature 2015?).]

If we cannot count on our memories about why we made such and such a design or analysis decision, then what should we do? How can we minimize regret with our past decisions? How can we improve our relationship with our future self?  This essay is a heavily revised and updated version of @bowers2011tpm and provides some suggestions for practices that will make such reproduction occur much more easily and quickly in the event that famous papers require special scrutiny.  Specifically, this piece aims to amplify some of what we already ought to know, and to add to some of those ideas given current practices, platforms, and possibilities.^[@king1995replication and @nagler1995coding were two of the first pieces introducing these kinds of ideas to political scientists. Now, the efforts to encourage transparency and ease of learning from the data and analyses of others has become institutionalized with the DA-RT inititiave <http://www.dartstatement.org/> @lupia2014openness. These ideas are spreading beyond political science as well @freese2007replication and @asendorpf2013recommendations.] This essay is an attempt at providing practical advice about how to *do work* such that one complies with such recommendations as a matter of course and, in so doing, can focus personal regret on other bad past decisions but not on decisions about data analysis and the production of scholarly papers.

@@

More importantly, we and others want to know why we did what we
did. Such explanations may not be very clear if we have some pages of
printed code in one hand and a manuscript in the other. Keep in mind
the distinction between the 'source code' of a document (i.e. what
computation was required to produce it) and the visible, type-set page
image. Page images are great for reading, but not great for
reproducing or collaborating. The source code of any document
exchanged by the group must be available and executable.^[As of the 2016 version of this paper, this idea is now widespread and made much easier than before via online services for code sharing and collaboration such as [GitHub](http://github.com) and [BitBucket](http://bitbucket.org).]
[I FEEL WE CAN MOVE THIS TO THE MAIN TEXT AND (BRIEFLY) INTRODUCE GITHUB]

@@

For example, in [@sec:data-analys-comp], I suggested that we know
where 'that plot on page 10 'comes from by making sure we had a
`fig1.pdf` file produced from a clearly commented plain text file
called something like `makefig1.R`.  An even easier solution would be to
directly include a chunk of code to produce the figure inside of the
paper itself. This paper, for example, was written in plain text using
markdown markup with R code chunks to make things like
Figure \ref{fig:giniprot}. This combination of markdown and R is called
[R Markdown](http://rmarkdown.rstudio.org).

@@

Markdown (and LaTeX and HTML) all have ways to cross-reference within a
document. For example, by using `\label{fig:giniprot}` in LaTeX or
`{#fig:giniprot}` in the pandoc version of markdown, I do not need to keep
track of the figure number, nor do extra work when I reorganize the document in
response to reviewer suggestions. Nor do I need a separate `fig1.R`file or
`fig1.pdf` file.  Tables and other numerical results are also possible to
generate within the source code of a scholarly paper. Those who view the code
for this essay will see how Table \ref{tab:protest} was also generated directly
from a regression object.\footnote{@beck2010reg inspired this particular
presentation of a linear model.}

@@

Group work requires version control.^[@fredrickson2011tpm and
@healy2011tpm in this issue also explain what version control is and why we
might want to use it.] Many people are familiar with the track changes
feature in modern WYSIWYG word processors or the fact that Dropbox allows one
to recover previous versions of files. These are both kinds of version control.
More generally, when we collaborate, we'd like to do a variety of actions with
our shared files. Collaboration on data analytic projects is more productive
and better when (1) it is easy to see what has changed between versions of
files; (2) members of the team feel free to experiment and then to dump parts
of the experiment in favor of previous work while merging the successful parts
of the experiment into the main body of the paper; (3) the team can produce
have 'releases' of the same document (one to MPSA, one to APSR, one to your
parents) without spawning many possibly conflicting copies of the same
document; (4) people can work on the same files at the same time without
conflicting with one another (and can reconcile their changes without too much
confusion and clobbering).  Clobbering is what happens when your future self or
your current collaborator saves an old version of a file over a new version,
erasing good work by accident.

Of course if you rely on Dropbox or 'track changes' in Word for version
control, you must communicate with other folks in your group before you edit
existing files. Only one of you can edit and save a given file at a time. This
prevents your work (or your colleagues work) from getting lost when you both
try to save the same file on top of each other. If you find that you need to
work on the same files at the same time, then you should work on establishing
your own shared version control system. Free options include launchpad, github,
sourceforge for open source projects (i.e. papers you are writing which you are
happy to share with others as you write). As of this writing, Google Docs does
an excellent job with shared editing of files as long as your are online. Each
of those services include paid versions too. One may also use Dropbox as a kind
of server for version control: for example, one may copy files from the Dropbox
directory into a local working directory so as to avoid clobbering and then
work on merging changes by hand before copying back to the Dropbox directory
and replacing existing files. MORE PERHAPS ON GOOGLE? LESS ON DROPBOX?

We use git with our own research group, and I use it for all of
my own projects. MORE ON GITHUB

Of course, one may take advantage of many of the benefits of formal
version control systems with some disciplined systems for file and
directory organization.  An excellent, simple, and robust version
control system is to rename your files with the date and time of
saving them: thedoc.tex becomes thedoc25-12-2011-23:50.tex.  Be sure
to include year in the file names --- remember, the life of an idea is
measured in years. If you are wise enough to have saved your documents
as plain text then you can easily compare documents using the many
utilities available for comparing text files.^[Adobe Acrobat
  allows one to compare differences in pdf files. OpenOffice supports
  a 'Compare Documents 'option. And Google Docs will report on the
  version history of a document.] When you reach certain milestones
you can rename the file accordingly: thedocAPSA2009.tex --- for the
one sent to discussants at APSA --- or thedocAPSR2015.tex --- for the
version eventually sent to the APSR six years after you presented it
at APSA. The formal version control systems I mentioned above all
allow this kind of thing and are much more elegant and capable, but
you can do it by hand too as long as you don't mind taking up a lot of
disk space and having many 'thedoc... ' files around. If you do
version control by hand, spend a little extra time to ensure that you
do not clobber files when you make mistakes typing in the
file-names. And, if you find yourself spending extra time reconciling
changes made by different collaborators by hand, remember this is a
task that modern version control systems take care of quickly and
easily.

@@

[IS THIS SECTION MORE MORE ON CREATING AN INTEGRATED REPRODUCABLE WORKFLOW RATHER THAN COPYING AND IMPROVING?]
Lots of people are thinking about 'reproducible research', creating a 'reproducable workflow' and
'literate programming' these days. Google those terms. Of course, the
devil is in the details: Here I list a few of my own attempts at enabling
reproducible research. You'll find many other inspiring examples on
the web. Luckily, the open source ethos aligns nicely with academic
incentives, so we are beginning to find more and more people offering
their files online for copying and improvement. By the way, if you do
copy and improve, it is polite to alert the person from whom you made
the copy about your work.

I have experimented with three systems so far: (1) for one paper we simply
included a Sweave document and data files into a compressed archive
[@bowers2005dataverse]; (2) for another more computing intensive paper we
assembled a set of files that enabled reproduction of our results using the
`make` system [@bowers2008dataverse]; (3) I have tried the "compendium"
approach [@gentleman2005reproducible,@gentleman2007statistical] which embeds an
academic paper within the R package system [@bowers2011dataverse]; and (4)
recently I have started organizing my work on github which enables me to make
"releases" of the project which should enable smooth reproduction of all of the
products of the research (simulations, data analyses, tables, figures, etc..).
The benefit of the compendium approach is that one is not required to have
access to a command line for `make`: the compendium is downloadable from within
R using `install.packages()` and is viewable using the `vignette()` function in
any operating system than runs R.^[Notice that my reproduction archives and/or
instructions for using them are hosted on the
[Dataverse](http://thedata.org/book/learn-about-project), which is another
system designed to enhance academic collaboration across time and space.]  The
idea that one ought to be able to install and run and use an academic paper
just as one installs and uses statistical software packages is very attractive,
and I anticipate that it will become ever easier to turn papers into R packages
as creative and energetic folks turn their attention to the question of
reproducible research.^[See for example <http://r-pkgs.had.co.nz/>]

[ADD OPEN SCIENCE FRAMEWORK, COLLABORATING ON DROPBOX AND NEW INITIATIVES LIKE Dynamic Documents (R Markdown/KnitR in R Studio,
Markdoc in Stata), Jupyter Notebook, Docker for Cross-Platform Reproducibility)]

**Step 6** We all learn by doing. When we create a reproducible workflow and share
reproduction materials we improve both cumulation of knowledge and our
methods for doing social science
[@freese2007replication,king1995replication}.  As we copy and
improve upon each other's modes of doing work we enhance our collective
ability to believe each other and for future scholars to believe us, too.
[I FEEL LIKE THE COPY COMMENT IS PERHAPS ANOTHER POINT TO MAKE (OR LEAVE)]

>>>>

---
title: How to improve your relationship with your future self.
author:
- name: Jake Bowers
  affiliation: <jwbowers@illinois.edu>,  University of Illinois at Urbana-Champaign, Departments of Political Science and Statistics; Fellow, White House Social and Behavioral Sciences Team. Nothing in this paper reflect the official position of the US Government. Many thanks to Mark Fredrickson, Brian Gaines, Kieran Healy, Kevin Quinn and Cara Wong for direct help on this document and to Mika LaVaque-Manty and Ben Hansen for many useful discussions on this topic. The source code for this document may be freely downloaded and modified from <https://github.com/jwbowers/workflow>.
- name: Maarten Voors
  affiliation: <maarten.voors@wur.nl>, Wageningen University, Development Economics Group. Special thanks to the BITSS team and the Department of Economics and Ted Miguel at UC Berkeley where Maarten was a visiting researcher during 2016.
bibliography: references.bib
csl: american-political-science-review.csl
date: '`r format(Sys.Date(), "%B %d, %Y")`'
output:
    pdf_document:
      keep_tex: true
      number_sections: true
      fig_width: 7
      fig_height: 7
      fig_caption: true
      template: bowersarticletemplate.latex
      pandoc_args: [
      "-M", "secPrefix=\\S",
      "--filter", "/usr/local/bin/pandoc-crossref"
      ]
      md_extensions: +autolink_bare_uris+ascii_identifiers+tex_math_single_backslash
...

@@

Memory is tricky. When we really want to remember something we should practice it until it becomes internalized. In fact, learning requires effort. And when we do not practice, we tend to be overconfident in our recall abilities [@koriat2005illusions] and false memories can be implanted.^[See the following site for a nice overview of what we know about memory including the fact that learning requires practice: <http://www.spring.org.uk/2012/10/how-memory-works-10-things-most-people-get-wrong.php>. On false memory see Wikipedia and linked studies <https://en.wikipedia.org/wiki/False_memory>.] If we cannot count on memory, then how can we do science?

How long does it take from planning a study to publication? And subsequently, how much time passes to the first reproduction of a study? Is three years too short? Is ten years too long? We suspect that few of our colleagues in the social and behavioral sciences conceive of, field, write, and publish a data driven study faster than about three years. We also suspect that, if some other scholar decides to re-examine the analyses of a published study, it will occur after publication. Moreover, this new scholarly activity of learning from one anothers' data and analyses, can occur at anytime, years past the initial publication of the article.^[We note that sometimes this reproduction occurs within the context of classes and learning about statistics and data analysis. Other times, this process is organized more broadly (cite IIIE program, cite Open Science Foundation on psychology, cite the Camerer 2016 Science paper?, cite the cool "Crowdsourcing Analysis" project by Silberzahn and Uhlmann Nature 2015?).]

If we cannot count on our memories about why we made such and such a design or analysis decision, then what should we do? How can we minimize regret with our past decisions? How can we improve our relationship with our future self?  This essay is a heavily revised and updated version of @bowers2011tpm and provides some suggestions for practices that will make such reproduction occur much more easily and quickly.  Specifically, this piece aims to amplify some of what we already ought to know, and to add to some of those ideas given current practices, platforms, and possibilities.^[@king1995replication and @nagler1995coding were two of the first pieces introducing these kinds of ideas to political scientists. Now, the efforts to encourage transparency and ease of learning from the data and analyses of others has become institutionalized with the DA-RT inititiave <http://www.dartstatement.org/> @lupia2014openness. These ideas are spreading beyond political science as well @freese2007replication and @asendorpf2013recommendations.] This essay is an attempt at providing practical advice about how to *do work* such that one complies with such recommendations as a matter of course and, in so doing, can focus personal regret on other bad past decisions but not on decisions about data analysis and the production of scholarly papers.

@@

More importantly, we and others want to know why we did what we
did. Such explanations may not be very clear if we have some pages of
printed code in one hand and a manuscript in the other. Keep in mind
the distinction between the 'source code' of a document (i.e. what
computation was required to produce it) and the visible, type-set page
image. Page images are great for reading, but not great for
reproducing or collaborating. The source code of any document
exchanged by the group must be available and executable.^[As of the 2016 version of this paper, this idea is now widespread and made much easier than before via online services for code sharing and collaboration such as [GitHub](http://github.com), [Open Science Framework](https://osf.io/) and [BitBucket](http://bitbucket.org), see below.]

@@

For example, in [@sec:data-analys-comp], we suggested that we know
where 'that plot on page 10' comes from by making sure we had a
`fig1.pdf` file produced from a clearly commented plain text file
called something like `makefig1.R`.  An even easier solution would be to
directly include a chunk of code to produce the figure inside of the
paper itself. This paper, for example, was written in plain text using
markdown markup with R code chunks to make things like
Figure \ref{fig:giniprot}. This combination of markdown and R is called
[R Markdown](http://rmarkdown.rstudio.org).

@@

Markdown (and LaTeX and HTML) all have ways to cross-reference within a
document. For example, by using `\label{fig:giniprot}` in LaTeX or
`{#fig:giniprot}` in the pandoc version of markdown, I do not need to keep
track of the figure number, nor do extra work when I reorganize the document in
response to reviewer suggestions. Nor do I need a separate `makefig1.R`file or
`fig1.pdf` file.  Tables and other numerical results are also possible to
generate within the source code of a scholarly paper. Those who view the code
for this essay will see how Table \ref{tab:protest} was also generated directly
from a regression object.\footnote{@beck2010reg inspired this particular
presentation of a linear model.}

@@

Group work requires version control. Many people are familiar with the track changes
feature in modern WYSIWYG (what you see is what you get) word processors or the fact that Dropbox allows one
to recover previous versions of files. These are both kinds of version control.
More generally, when we collaborate, we'd like to do a variety of actions with
our shared files. Collaboration on data analytic projects is more productive
and better when (1) it is easy to see what has changed between versions of
files; (2) members of the team feel free to experiment and then to dump parts
of the experimentation in favor of previous work while merging the successful parts
 into the main body of the paper; (3) the team can produce
have 'releases' of the same document (one to MPSA, one to APSR, one to your
parents) without spawning many possibly conflicting copies of the same
document; (4) people can work on the same files at the same time without
conflicting with one another, and can reconcile their changes without too much
confusion and clobbering. Clobbering is what happens when your future self or
your current collaborator saves an old version of a file over a new version,
erasing good work by accident.

Of course if you rely on Dropbox or 'track changes' in Word for version
control, you must communicate with other folks in your group before you edit
existing files. Only one of you can edit and save a given file at a time. This
prevents your work (or your colleagues work) from getting lost when you both
try to save the same file on top of each other. If you find that you need to
work on the same files at the same time, then you should work on establishing
your own shared version control system. Free options include launchpad, github, Open Science Framework,
sourceforge for open source projects (i.e. papers you are writing which you are
happy to share with others as you write, see more below). Each has a slightly difference language (some say folders other repositories, some say download othere clone, etc) and it takes a bit of time adjusting. But as we'll argue: it is well worth the time investment as it will save (lots!) of time later on. Currently Google Docs does
an excellent job with shared editing of files. The downside is: you have to be online. For people with irregular or slow internet connections, this may not be an option. MORE ON GOOGLE?
One may also use Dropbox as a kind of server for version control (though the downside is if your collaborations or project grows large (in terms of MBs) you need to switch to a paid service). Maarten has been using this for most of his collaborations. You can copy files from the Dropbox
directory into a local working directory so as to avoid clobbering and then
work on merging changes by hand before copying back to the Dropbox directory
and replacing existing files. You will however need to agree on an iron file and directory naming rule with your collaborators.

Jake uses github within his research group, and used it for all of
his own projects. MORE ON GITHUB
You may even obtain a two year unlimited repositories with an academic affiliation.

 An excellent, simple, and robust version
control system is to rename your files with the date and time of
saving them: yyyymmdd_project.docx (for changes within the same day, just before a deadline for example, you may even add a time, so it becomes 20160623_4PM_inequalitypaper.docx).  Be sure to include year in the file names --- remember, the life of an idea is
measured in years. If you are wise enough to have saved your documents
as plain text then you can easily compare documents using the many
utilities available for comparing text files.^[Adobe Acrobat
  allows one to compare differences in pdf files. OpenOffice supports
  a 'Compare Documents 'option. Word now does the same. And Google Docs will report on the
  version history of a document.] When you reach certain milestones
you can rename the file accordingly: thedocAPSA2009.tex --- for the
one sent to discussants at APSA --- or thedocAPSR2015.tex --- for the
version eventually sent to the APSR six years after you presented it
at APSA. The formal version control systems mentioned above all
allow this kind of thing and are much more elegant and capable, but
you can do it by hand too as long as you don't mind taking up a lot of
disk space and having many 'thedoc... ' files around. Indeed, the number of files can add up quickly (especially around submission time!) and you may want to create an "0_Archive" folder to store older versions. If you do
version control by hand, spend a little extra time to ensure that you
do not clobber files when you make mistakes typing in the
file-names. And, if you find yourself spending extra time reconciling
changes made by different collaborators by hand, remember this is a
task that modern version control systems take care of quickly and
easily.

@@

[IS THIS SECTION MORE MORE ON CREATING AN INTEGRATED REPRODUCIBLE WORKFLOW RATHER THAN COPYING AND IMPROVING?]
Lots of people are thinking about 'reproducible research', creating a 'reproducible workflow' and
'literate programming' these days. Google those terms. Of course, the
devil is in the details: Here I list a few of my own attempts at enabling
reproducible research. You'll find many other inspiring examples on
the web. Luckily, the open source ethos aligns nicely with academic
incentives, so we are beginning to find more and more people offering
their files online for copying and improvement. By the way, if you do
copy and improve, it is polite to alert the person from whom you made
the copy about your work.

We have experimented with several systems so far: (1) we wrote this paper in R markdown (Jake) and Atom (Maarten) and organized our files and collaboration on Github. The source code can be accessed, downloaded and modified from <https://github.com/jwbowers/workflow>. The program also allowed us to send notes to each other. The nice thing about github is that is enables you to make "releases" of a project which enable you to smooth the reproduction of all of the
products of your research (simulations, data analyses, tables, figures, etc..). Similar features are offered by the Open Science Framework (https://osf.io/), Maarten recently participated in a workshop on research transparency organized by the Berkeley Institute for Transparency in the Social Sciences (www.bitss.org), all course material was accessible through OSF (https://osf.io/qh2nr/). Both platforms, OSF and Github nicely integrate and a key benefit is that they have been created to remain for a long time (unlike perhaps your university drive); (2) for one paper Jake simply included a Sweave document and data files into a compressed archive [@bowers2005dataverse] [SAY ABIT MORE ABOUT WHAT SWEAVE IS IN PRINCIPLE]; (3) for another more computing intensive paper Jake
assembled a set of files that enabled reproduction of our results using the
`make` system [@bowers2008dataverse] [EXPLAIN A BIT MORE OR INCLUDE REF TO WEBSITE?]; (4) Jake also tried the "compendium"
approach [@gentleman2005reproducible,@gentleman2007statistical] which embeds an
academic paper within the R package system [@bowers2011dataverse]; The benefit of the compendium approach is that one is not required to have
access to a command line for `make`: the compendium is downloadable from within
R using `install.packages()` and is viewable using the `vignette()` function in
any operating system than runs R.^[Notice that Jake's reproduction archives and/or
instructions for using them are hosted on the
[Dataverse](http://thedata.org/book/learn-about-project), which is another
system designed to enhance academic collaboration across time and space.]  The
idea that one ought to be able to install and run and use an academic paper
just as one installs and uses statistical software packages is very attractive,
and we anticipate that it will become ever easier to turn papers into R packages
as creative and energetic folks turn their attention to the question of
reproducible research.^[See for example <http://r-pkgs.had.co.nz/>]
Finally, (5) Maarten has been using Dropbox for most of his collaborative projects (https://www.dropbox.com/). ALREADY ABOVE?

There are some new noteworthy developments also. For example, and similar to R Markdown/KnitR in R Studio, there is Markdoc for STATA users (https://github.com/haghish/MarkDoc). People at UC Berkeley have been developing a is a web app, Jupyter Notebook (http://jupyter.org/), that enables you to make and share documents that include text, code, equations and visualizations in one. For researchers working across difference platforms, there is Docker (https://www.docker.com/).

**Step 6** We all learn by doing. When we create a reproducible workflow and share
reproduction materials we improve both cumulation of knowledge and our
methods for doing social science
[@freese2007replication,king1995replication}.

 As we copy and
improve upon each other's modes of doing work we enhance our collective
ability to believe each other and for future scholars to believe us, too.
[I FEEL LIKE THE COPY COMMENT IS PERHAPS ANOTHER POINT TO MAKE (OR LEAVE)]
